---
title: "HOMEWORK 1 - Proportion Not Returned"
author: "Rui Li"
date: "8/20/2020"
header-includes:
- \usepackage{amsgen,amsmath,amstext,amsbsy,amsopn,amssymb,mathabx,amsthm,bm,bbm}
- \usepackage[labelsep=space]{caption}
output:
  word_document: default
  html_document: default
subtitle: $\textbf{STOR 590, FALL 2020}$
---

# Import and Summarize the Dataset
```{r, eval = TRUE}
#Import Proportion Not Returned.csv
library(readr)
dataset <- read_csv("ProportionNotReturned.csv")

#Have a general idea about the dataset
head(dataset)
print(summary(dataset))
```

# Find the optimal model to predict PNR
```{r, eval = TRUE}
#Summarize the PNR variable
library(ggplot2)
ggplot(dataset, aes(x=PNR)) + 
  geom_histogram(aes(y=..density..), bins = 20,fill = "white", col = "black") + 
  geom_density(alpha=.2, fill="#FF6666") +
  geom_rug() +
  labs(title = 'Proportion Absentee Ballots Not Returned - NC Nov 2018') + 
  theme(plot.title = element_text(hjust = 0.5, size=12, face="bold.italic"))

#Omitting Bladen and Robeson counties with weights
wts = dataset$PNR<0.1
#Extract PNR and other predicted variables
pnr.df = dataset[,2:11]
#Build up full model
full.lm = lm(formula = PNR ~ ., data = pnr.df, weights=as.numeric(wts))
print(summary(full.lm))

#Apply backward selection model
full.backward = step(full.lm, direction = "backward")
print(summary(full.backward))

#Apply forward selection model
full.forward <- step(lm(PNR ~ 1, data=pnr.df), list(upper=full.lm), direction='forward')
print(summary(full.forward))
```

I apply both forward selection and backward selection to the full model, and get two different optimal models. The optimal model of backward selection is *PNR ~ Pop + Hsgrad + Black*, while the one of forward selection is *PNR ~ Black*.

In the following step, I will use diagnostics to assess various measures of fit and choose the better model.
```{r, eval = TRUE}
#Optimal model of backward selection
backward.lm = lm(formula = PNR ~ Pop + Hsgrad + Black, data = pnr.df, weights = as.numeric(wts))
#Diagnostics of backward optimal model
plot(backward.lm)

#Optimal model of forward selection
forward.lm = lm(formula = PNR ~ Black, data = pnr.df, weights = as.numeric(wts))
#Diagnostics of backward optimal model
plot(forward.lm)
```

**Diagnosis Results:** From the *Residuals vs. fitted values* plots of both models, we can see that both scatterplots distribute randomly, with some exception outliers in the bottom right corner, but the backward model's is more haphazard.

From the *Normal Probability plot* of both models, the errors of the backward model is more distributed normally, with a better straight line than the forward model. 

From the *Scale Lotion* plots, both models show radom patterns. 

From the *Cook's distance* plots, we can see there are some outliers affecting both models. 

Thus, according to the results above, I will pick the backward model,*PNR ~ Pop + Hsgrad + Black*, as my optimal model for the future steps. 

# PNR prediction interval for Bladen and Robeson counties
```{r, eval = TRUE}
#Choose backward model as the optimal model
opt.lm = backward.lm
#Predict based on 99% prediction interval
opt.pre.99=predict(opt.lm,se.fit=T,interval='prediction',level=0.99,weights=1)
opt.pre.99$fit[c(9,78),]
```

The 99% prediction interval of Bladen and Robeson are *(0.00058, 0.03399)*, and *(-0.00131, 0.03242)*.

# Estimate excess PNR for Bladen and Robeson counties
```{r, eval = TRUE}
excess.PNR=pnr.df[c(9,78),'PNR']-opt.pre.99$fit[c(9,78),'upr']
excess.PNR
```

The excee PNR of Bladen and Robeson are *0.079*, and *0.078* resepectively. 

# Estimate the total number of absentee ballots that are unaccounted for
```{r, eval = TRUE}
total=sum(excess.PNR*dataset[c(9,78),'AbsBal'])
print(total)
```

The total number of absentee ballots that are unaccounted for is 1888.236.

**Question** The actual number of votes by which Mark Harris was leading at the time the count was stopped was 905. The Harris campaign responded to the allegations by asserting that the number of potentially missing votes was very small and certainly less than 905. Does your analysis support that conclusion - why or why not?
```{r}
print(905/(905+total))
```

**Answer** Compared to the predicted missing votes of 1888, 905 only represents a minor group of people. According to my analysis, the valid votes only represent 32.40% of the whole. Thus, my analysis does not support Harris's conclusion. 